Live Functional Programming with Typed Holes

### 2018-04-02 calculus.tex
draft of 3.3

Non-empty expression holes, written $\hhole{\hexp}{u}$, behave similarly. Rule {SNEHole} specifies that non-empty expression holes also synthesize hole type as long as the expression inside the hole, $\hexp$, synthesizes some (arbitrary) type. Non-empty expression holes therefore internalize the ``red squiggles'' that many editors display under or around type inconsistencies in a program.\todo{example?}

@@

\subsubsection{Hole Expansion} Rules {ESEHole}, {ESNEHole}, {EAEHole} and {EANEHole} govern the expansion of empty and non-empty expression holes to \emph{hole closures}, $\dehole{u}{\sigma}{}$ and $\dhole{\dexp}{u}{\sigma}{}$, respectively.

@@

The hole expansion rules are the only rules that introduce hypotheses, of the form $\Dbinding{u}{\hGamma}{\htau}$, into the hole context, $\Delta$. The purpose of the hole context is to record a typing context, $\Gamma$, and a type, $\tau$, for each hole name, $u$.\footnote{We use a hole context, rather than recording the typing context and type directly on each hole closure, to ensure that all closures for a hole name have the same typing context and type.} This notation for hole contexts is taken from contextual modal type theory (CMTT) \cite{Nanevski2008}, identifying hole names with metavariables and hole contexts with metavariable contexts (we say more about the connection with CMTT below). In all four hole expansion rules, the typing context recorded in the hole context is simply the current typing context when the hole is expanded. In the synthetic hole expansion rules, {ESEHole} and {ESNEHole}, the generated hole context assigns the hole type, $\tehole$, to $u$, as in the typing rules. However, the first two premises of the expansion subsumption rule, Rule EASubsume, disallow the use of subsumption for holes in analytic position during expansion. Instead, we have separate analytic rules, {EAEHole} and {EANEHole}, which record the type that the hole is being checked against into the hole context. This is again so that we can use a type assignment system for the internal language --- the type assignment rules TAEHole and TANEHole in Fig.~\ref{fig:hasType} assign a hole closure for $u$ the corresponding type from the hole context.

Each hole closure also has an associated environment, $\sigma$, which is a finite substitution of the form $[d_1/x_1, ~\cdots, d_n/x_n]$ for $n \geq 0$. The purpose of the closure environment is to keep a record of the substitutions that occur around the hole as evaluation occurs. Initially, no evaluation has occured, so the initial environment generated by the hole expansion rules is the identity substitution for the typing context associated with $u$ in $\Delta$, which we notate $\idof{\hGamma}$ and define as follows.
\begin{defn}[Identity Substitution] $\idof{x_1 : \tau_1, ~\cdots, x_n : \tau_n} = [x_1/x_1, ~\cdots, x_n/x_n]$
\end{defn}
\noindent
The type assignment rules for hole closures, Rules TAEHole and TANEHole, require that we be able to check the environment of each hole closure against the corresponding typing context, written $\hasType{\Delta}{\hGamma}{\sigma}{\hGamma'}$ and defined as follows:\todo{check that the definition in the Agda corresponds}
\begin{defn}[Substitution Typing]
$\hasType{\Delta}{\hGamma}{\sigma}{\hGamma'}$ iff $\domof{\sigma} = \domof{\hGamma'}$ and for each $x : \htau \in \hGamma'$ we have that $d/x \in \sigma$ and $\hasType{\Delta}{\hGamma}{d}{\tau}$.
\end{defn}
\noindent
It is easy to verify that the identity substitution satisfies this requirement, i.e. that $\hasType{\Delta}{\hGamma}{\idof{\hGamma}}{\hGamma}$. We will see how closure environments evolve during evaluation in Sec.~\ref{sec:evaluation}.

\subsubsection{Cast Insertion} 
Consider the following example: $\hap{(\halam{x}{\tehole}{\hap{x}{c}})}{c}$. The type synthesized for this example viewed as an external expression is $\tehole$, because the hole type annotation on $x$ allows us to apply it as a function of type $\tarr{\tehole}{\tehole}$ as previously discussed, and the outer argument, $c$, can be checked against type $\tehole$ by subsumption. However, viewed as an internal expression, this example is not well-typed---we do not have subsumption in the type assignment system defined in Fig.~\ref{fig:hasType}. Indeed, it would violate type safety if we could assign a type to this example in the internal language, because beta reduction of this example viewed as an internal expression would result in $c(c)$, which is clearly not well-typed ($c$ does not have arrow type). The problem is that leaving the argument type unknown leaves how the argument is being used (in this case, as a function) also unknown.\footnote{In a system where type reconstruction is first used to try to fill in type holes, we could express a similar example by using $x$ at two or more different types, thereby causing type reconstruction to fail.
%On the other hand, if it is acceptable to arbitrarily choose one of the possible types, and type reconstruction is complete, then type holes will never appear in the internal language and the cast insertion machinery described in this section can be omitted entirely, leaving only the hole closure machinery described previously.
}\todo{talk about elsewhere? maybe do type-hole-free version of calculus in appendix if there is time}{} Recall that we can interpret hole types as unknown types from gradual type theory. This leads us to the solution to this problem: cast insertion.

@@

\subsection{Dynamic Semantics}
\label{sec:evaluation}

@@

\begin{theorem}[Canonical Value Forms]
  If $\hasType{\hDelta}{\emptyset}{\dexp}{\htau}$ and $\isValue{\dexp}$
  then $\htau\neq\tehole$ and
  \begin{enumerate}[label=(\alph*)]
    \item If $\htau=b$ then $\dexp=c$.
    \item If $\htau=\tarr{\htau_1}{\htau_2}$
          then $\dexp=\halam{x}{\htau_1}{\dexp'}$
          where $\hasType{\hDelta}{x : \htau_1}{\dexp'}{\htau_2}$.
  \end{enumerate}
\end{theorem}

@@

\autoref{fig:complete} defines \emph{complete} types, expressions, and
dynamic expressions to be those terms without type or expression holes.

@@

%% \begin{figure}[!ht]
  \begin{definition}[Substitution Typing]
    $\hasType{\Delta}{\hGamma}{\sigma}{\hGamma'}$ iff $\domof{\sigma} = \domof{\Gamma'}$ and for each $\dexp/x \in \sigma$, we have $x : \htau \in \hGamma'$ and $\hasType{\Delta}{\hGamma}{\dexp}{\htau}$.
  \end{definition}
%%   \caption{substitution type assignment}
%% \end{figure}

>>>>

Non-empty expression holes, written $\hhole{\hexp}{u}$, behave similarly. Rule {SNEHole} specifies that non-empty expression holes also synthesize hole type as long as the expression inside the hole, $\hexp$, synthesizes some (arbitrary) type. Non-empty expression holes therefore internalize the ``red squiggles'' that many editors display under or around type inconsistencies in a program.\todo{example?}\todo{briefly say something about binding inconsistencies?}

@@

\subsubsection{Hole Expansion}\label{sec:hole-expansion} Rules {ESEHole}, {ESNEHole}, {EAEHole} and {EANEHole} govern the expansion of empty and non-empty expression holes to \emph{hole closures}, $\dehole{u}{\sigma}{}$ and $\dhole{\dexp}{u}{\sigma}{}$, respectively.

@@

The hole expansion rules are the only rules that introduce hypotheses, of the form $\Dbinding{u}{\hGamma}{\htau}$, into the hole context, $\Delta$. The purpose of the hole context is to record a type, $\tau$, and a typing context, $\Gamma$,   for each hole name, $u$.\footnote{We use a hole context, rather than recording the typing context and type directly on each hole closure, to ensure that all closures for a hole name have the same typing context and type.} This notation for hole contexts is taken from contextual modal type theory (CMTT) \cite{Nanevski2008}, identifying hole names with metavariables and hole contexts with modal contexts (we say more about the connection with CMTT below). In all four hole expansion rules, the typing context recorded in the hole context is simply the current typing context when the hole is expanded. In the synthetic hole expansion rules, {ESEHole} and {ESNEHole}, the generated hole context assigns the hole type, $\tehole$, to $u$, as in the typing rules. However, the first two premises of the expansion subsumption rule, Rule EASubsume, disallow the use of subsumption for holes in analytic position. Instead, we have separate analytic rules, {EAEHole} and {EANEHole}, which record the type that the hole is being checked against into the hole context. This is again so that we can use type assignment for the internal language --- the type assignment rules TAEHole and TANEHole in Fig.~\ref{fig:hasType} assign a hole closure for $u$ the corresponding type from the hole context.

Each hole closure also has an associated environment, $\sigma$, which is a finite substitution of the form $[d_1/x_1, ~\cdots, d_n/x_n]$ for $n \geq 0$. The purpose of the closure environment is to keep a record of the substitutions that occur around the hole as evaluation occurs. Initially, no evaluation has occured, so the initial environment generated by the hole expansion rules is the identity substitution for the typing context associated with $u$ in $\Delta$, which we notate $\idof{\hGamma}$ and define as follows.
\begin{defn}[Identity Substitution] $\idof{x_1 : \tau_1, ~\cdots, x_n : \tau_n} = [x_1/x_1, ~\cdots, x_n/x_n]$
\end{defn}
\noindent
The type assignment rules for hole closures, Rules TAEHole and TANEHole, require that we be able to check the environment of each hole closure against the corresponding typing context, written $\hasType{\Delta}{\hGamma}{\sigma}{\hGamma'}$ and defined as follows:\todo{check that the definition in the Agda corresponds}
\begin{defn}[Substitution Typing]
$\hasType{\Delta}{\hGamma}{\sigma}{\hGamma'}$ iff $\domof{\sigma} = \domof{\hGamma'}$ and for each $x : \htau \in \hGamma'$ we have that $d/x \in \sigma$ and $\hasType{\Delta}{\hGamma}{d}{\tau}$.
\end{defn}
\noindent
It is easy to verify that the identity substitution satisfies this requirement, i.e. that $\hasType{\Delta}{\hGamma}{\idof{\hGamma}}{\hGamma}$.

Empty hole closures, $\dehole{u}{\sigma}{}$,  correspond to the metavariable closures (a.k.a. deferred substitutions) from CMTT, $\cmttclo{u}{\sigma}$.\todo{change notation from CMTT}{} We will see how closure environments evolve during evaluation in Sec.~\ref{sec:evaluation}. Non-empty hole closures, $\dhole{d}{u}{\sigma}{}$, do not directly correspond to a notion from CMTT\todo{mention reason in Sec 4}.

\subsubsection{Cast Insertion}\label{sec:cast-insertion} 
Consider the following example: $\hap{(\halam{x}{\tehole}{\hap{x}{c}})}{c}$. The type synthesized for this example viewed as an external expression is $\tehole$, because the hole type annotation on $x$ allows us to apply it as a function of type $\tarr{\tehole}{\tehole}$, as previously discussed, and the outer argument, $c$, can be checked against type $\tehole$ by subsumption. However, viewed as an internal expression, this example is not well-typed---we do not have subsumption in the type assignment system defined in Fig.~\ref{fig:hasType}. Indeed, it would violate type safety if we could assign a type to this example in the internal language, because beta reduction of this example viewed as an internal expression would result in $c(c)$, which is clearly not well-typed ($c$ does not have arrow type). The problem is that leaving the argument type unknown leaves how the argument is being used (in this case, as a function) also unknown.\footnote{In a system where type reconstruction is first used to try to fill in type holes, we could express a similar example by using $x$ at two or more different types, thereby causing type reconstruction to fail.
%On the other hand, if it is acceptable to arbitrarily choose one of the possible types, and type reconstruction is complete, then type holes will never appear in the internal language and the cast insertion machinery described in this section can be omitted entirely, leaving only the hole closure machinery described previously.
}\todo{talk about elsewhere? maybe do type-hole-free version of calculus in appendix if there is time}{} Recall that we can interpret hole types as unknown types from gradual type theory. This leads us to the solution to this problem: cast insertion.

@@

\subsection{Evaluation}
\label{sec:evaluation}

To recap, the result of expansion is a well-typed internal expression with hole closures and casts. As described in Sec.~\ref{sec:hole-expansion} and Sec.~\ref{sec:cast-insertion} respectively, empty hole closures correspond to metavariable closures from CMTT, with the metavariable free (i.e. tracked in $\Delta$) \cite{Nanevski2008} and casts correspond to casts from gradual type theory \cite{Siek06a,DBLP:conf/snapl/SiekVCB15}. However, the dynamic semantics for \HazelnutLive does not simply ``fall out'' from these observations.

The problem is first that \citet{Nanevski2008} defined only the logical reductions for CMTT, viewing it as a proof system for intuitionistic contextual modal logic via the propositions-as-types (Curry-Howard) correspondence. The paper therefore proved only a subject reduction property, which corresponds to type preservation for elimination transitions, generalized to open terms\todo{citation?}{}. This is not a full dynamic semantics, and in particular, there is no notion of progress, i.e. that well-typed terms cannot get ``stuck'' in an undefined state. In any case, a conventional dynamic semantics for CMTT would not be immediately relevant to our goal of evaluating incomplete programs because, by our interpretation of hole closures, we would need an operational semantics for terms with free metavariables. \citet{Nanevski2008} sketched an interpretation of CMTT into the simply-typed lambda calculus with sums under permutation conversion\footnote{Permutation conversions are necessary to encode the commutative reductions of CMTT, which in turn are necessary to prove a strong normalization property. These issues are not relevant in \HazelnutLive because, as in the gradually typed lambda calculus, type holes admit non-termination: we can express the Y combinator as $(\halam{x}{\tehole}{x(x)}) (\halam{x}{\tehole}{x(x)})$.}, which has been studied by \citet{DBLP:journals/iandc/Groote02}, but under this interpretation the same problem arises -- metavariables become variables of a function type, so we cannot rely on any standard notion of progress on closed terms.\todo{citations}{}% We also cannot rely on, for example, weak head normalization because \HazelnutLive admits non-termination (due to casts).\todo{citation}

Of course, we also need to integrate casts into the dynamic semantics. Fortunately, the operational semantics for the cast calculus from the gradually typed lambda calculus provides most, but not all, of the necessary machinery. Briefly, the subtlety is that in the cast calculus, the only irreducible terms of hole type are casts, but in $\HazelnutLive$, hole closures induce additional irreducible terms. We will return to the technical implications of this fact below. The second missing piece in existing work on casts is that, when cast failure occurs, evaluation aborts (i.e. blame propagates). Our goal, as discussed in Sec.~\ref{sec:failed-cast-example}, is to have cast failure spawn a membrane around the dynamic type error, much like a non-empty hole serves as a membrane around a static type error, allowing evaluation to safely continue past the cast failure when desired.

This section defines a reduction semantics for well-typed internal expressions.

Matthias Felleisen, Robert Hieb,
The revised report on the syntactic theories of sequential control and state,

@@

@@

\subsection{Evaluating Complete Terms}\label{sec:complete-terms}

\autoref{fig:complete} defines \emph{complete} types, expressions, and
dynamic expressions to be those terms without type or expression holes.
\begin{theorem}[Canonical Value Forms]
  If $\hasType{\hDelta}{\emptyset}{\dexp}{\htau}$ and $\isValue{\dexp}$
  then $\htau\neq\tehole$ and
  \begin{enumerate}[label=(\alph*)]
    \item If $\htau=b$ then $\dexp=c$.
    \item If $\htau=\tarr{\htau_1}{\htau_2}$
          then $\dexp=\halam{x}{\htau_1}{\dexp'}$
          where $\hasType{\hDelta}{x : \htau_1}{\dexp'}{\htau_2}$.
  \end{enumerate}
\end{theorem}

@@

%% \begin{figure}[!ht]
%%   \caption{substitution type assignment}
%% \end{figure}
